---
title: "Madagascar Plague Model"
author: "Mats Slik & Dennis Haandrikman"
date: June 15th, 2022
output:
  pdf_document:
    toc: true
    toc_depth: 2
---


# Introduction


# Methods


# Results

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)

# Required libraries
library("deSolve")
library("pander")
```
```{r data}
data <- read.csv("C:\\Users\\matsp\\Downloads\\plague2017-master\\plague2017-master\\test2.csv", col.names = c('Sb','Sy','Eb','Ey','Ib','Ip', 'tijd'))
# options(digits = 15)

# Run a test to see if the data imported correctly
pander(head(data))
```
In the table above, the imported data is visible.
```{r variables}
# TODO fixe logical bug

# Define the variables, parameters, state & time.
tau_p <- 8.893084e+00
tau_b <- 1.793090e+01

A <- 1.15
B <- 0.08
C <- 0.1


parameters <- c(p       = 6.459494e-05, # fraction susceptible to bite
                alpha   = 1.904753e-03, # rate of bites
                beta    = 2.233049e+00, # between-host infection
                gamma_b = 2.293953e-01, # E to I
                gamma_p = 2.863875e-01, # E to I
                delta_b = 2.626527e-01, # recovery + death
                delta_p = 3.413748e-01, # recovery + death
                epsilon = 3.204494e-02, # I_b to I_p
                kk      = 1.105443e+00, # Imperfect intervention
                N       = 25570895     # population size
                )

S_b_0 <- 1651
S_p_0 <- 25569244

state <- c(Sb = S_b_0, Sy = S_p_0, Eb = 0, Ey =  0, Ib = 1, Ip = 1)

times <- seq(0, 70 - 1, by = 0.1)

# Define the flea function for the flea population
flea <- function (t){
  x <- A + B * sin((pi/180.) * t) + C * cos((pi/180.) * t )

  return (x)
}
# Define the intervention functions of human intervention on the disease
intervention_p <- function (t){
  return (1 - 1/(as.numeric(parameters["kk"]) + exp(tau_p - t)))
}
intervention_b <- function (t){
  return (1 - 1/(as.numeric(parameters["kk"]) + exp(tau_b - t)))
}

```
```{r Deterministic model}
# Define the function for the model of the deterministic data runs
deterministic_model <- function(t, y, parms){
  with(as.list(c(parms, y, t)),{
    flea_t <- flea(t)
    intervention_b1 <- intervention_b(t)
    intervention_p1 <- intervention_p(t)

    delta.Sb <- -alpha * flea_t * intervention_b1 * Sb - -beta * intervention_p1 * Sb * Ip/N
    delta.Sy <- -beta * intervention_p1 * Sy * Ip/N
    delta.Eb <- alpha * flea_t * intervention_b1 * Sb - gamma_b * Eb
    delta.Ey <- beta * intervention_p1 * (Sb+Sy) * Ip/N - gamma_p * Ey
    delta.Ib <- gamma_b * Eb - epsilon * Ib - delta_b * Ib
    delta.Ip <- gamma_p * Ey + epsilon * Ib - delta_p * Ip
    print(delta.Sb)


        return(list(c(delta.Sb, delta.Sy, delta.Eb, delta.Ey, delta.Ib, delta.Ip)))
       })
}

# Run the ode model and save the output to the out variable
out <- ode(times = times, y = state,   parms = parameters, func = deterministic_model, method = "lsoda", atol = 1e-10, rtol = 1e-10)

plot(as.data.frame(out)$Eb, type = "l", xlab = "Time in Days", ylab = "Bubonice indices", col = "blue", ylim = c(0, 30))
plot(as.data.frame(out)$Ey, type = "l", xlab = "Time in Days", ylab = "Puneonic indices", col = "blue", ylim = c(0, 120))
plot(out)
plot(data$Ip)

```
In the graph(s), we can see the plot of the deterministic data model.
```{r stochastic}
# Vectorize the flea & intervention functions
# v_flea <- vectorize(flea)
# v_itv_p <- vectorize(intervention_p)
# v_itv_b <- vectorize(intervention_b)

# Save all the flea/intervention data based on each piece of time
f_irf <- flea(t)
f_itv_b <- intervention_b(t)
f_itv_p <- intervention_p(t)

unit_vector <- function (i, D){
  out <- rep(0, D)
  out[i] <- 1
  return (out)
}

# Get the time and make an index with said time
get_time <- function (rates, Pk, Tk){
  n <- length(rates)
  for (i in range(n)){
    if (rates[i] > 0.0){
      dt <- array(Pk[i] - Tk[i] / rates[i])
    } else {
      break
    }
  }
  idx <- argmin(dt)
  return (idx, dt[idx])
}

transition_rates <- function (f_irf, f_itv_p, f_itv_b) {
  return (array(parameters[["alpha"]] * f_irf * f_itv_b * state[["Sb"]],
                parameters[["beta"]] * f_itv_p * state[["Sb"]] * state[["Ip"]]/parameters[["N"]],
                parameters[["beta"]] * f_itv_p * state[["Sy"]] * state[["Ip"]]/parameters[["N"]],
                parameters[["gamma_b"]] * state[["Eb"]],
                parameters[["gamma_p"]] * state[["Ey"]],
                parameters[["delta_b"]] * state[["Ib"]],
                parameters[["delta_p"]] * state[["Ip"]],
                parameters[["epsilon"]] * state[["Ib"]])
  )
}

nu <- array(unit_vector(2, D) - unit_vector(0, D), unit_vector(3, D) - unit_vector(0, D),
               unit_vector(3, D) - unit_vector(1, D), unit_vector(4, D) - unit_vector(2, D),
               unit_vector(5, D) - unit_vector(3, D), -unit_vector(4, D), -unit_vector(5, D),
               unit_vector(5, D) - unit_vector(4, D), dtype=int)

change_func <- function(i, n, firings, nu) {
  prechange <- c()
  for (i in range(n)) {
    append(firings[i]*nu[i], prechange)
  }
  return( sum(prechange))
}

plague_tau <- function (parms, t, f_irf, f_itv_p, f_itv_b, nu) {
  x_t <- list()
  for (i in range(length(t))){
    append(x_t, x)
    rates <- transition_rates(x, f_irf[i], f_itv_p[i], f_itv_b[i])
    n <- length(rates)
    firings <- rpois(max(rates, rep(0, n))*dt)
    change <- change_func(i, n, firings, nu)
    x <- x + change
  }
  x_t <- array(x_t)
  return(x_t)
}

# Set the seed for the random number generator
set.seed(10042018)
runs <- 20

I_b <- c()
I_P <- c()

for (realisation in range(runs)){
  y <- plague_tau(parms, t, f_irf, f_itv_p, f_itv_b, nu)
  append(y[,4], I_b)
  append(y[,5], I_P)
}

```

# Conclusion


# Discussion

in translating the python model to R
We found that there are some minor internal rounding differences when calculating floats in R and python,
these differences where isolated and removed, simply by comparing each outcome of the r and python equation and.
the outcome of each equation used for calculating the delta value was exactly the same.
This let to that we found that DeSolve ode function and the scypy odeint have minor differences calculation the deltas with a given time that create a snow balling effect.
this results in the R modell having a higher peak curve compared to the Python 3 model.

# Sources
