---
title: "The 2017 plague outbreak in Madagascar & the Madagascar Plague Model"
author:
  - name: Mats Slik
    affiliation: Hanze University of Applied Sciences
    address:
    - Zernikeplein 11
    - room D1.12
    email: m.p.slik@st.hanze.nl
  - name: Dennis Haandrikman
    affiliation: Hanze University of Applied Sciences
    address:
    - Zernikeplein 11
    - room D1.12
    email: d.haandrikman@st.hanze.nl
abstract: >
  During this experiment there
date: June 15th, 2022
output:
    pdf_document:
      toc: true
      toc_depth: 2
---


# Introduction

The goal of this experiment was to learn more about the behaviour and the way the plague spreads and then modelling our findingss about the plague.
Due to the fact that Madagascar has annual plague outbreaks, it was decided to use their outbreak information/data as data for this experiment.

The plague has once been one of the deadliest natural disasters in human history, referred to as the Black Death, attributed to the bacterium 'Yersinia pestis'. (
Although the disease was widespread in ancient times, nowadays, plague epidemics still pose a threat to humans, as there are reports of continues annual infections in the following 5 countries:
Madagascar, Tanzania, Vietnam, China & the USA.

One of the goals set for this experiment is to recreate the model and results that the researchers gained from their experiments.
To remake the model, the formula's used in the document have been

# Methods


# Results

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)

# Required libraries
library("deSolve")
library("pander")
```
```{r data}
data <- read.csv("C:\\Users\\matsp\\Downloads\\plague2017-master\\plague2017-master\\test2.csv", col.names = c('Sb','Sy','Eb','Ey','Ib','Ip', 'tijd'))
# options(digits = 15)

# Run a test to see if the data imported correctly
pander(head(data))
```
In the table above, the imported data is visible.
```{r variables}
# TODO fixe logical bug

# Define the variables, parameters, state & time.
tau_p <- 8.893084e+00
tau_b <- 1.793090e+01

A <- 1.15
B <- 0.08
C <- 0.1


parameters <- c(p       = 6.459494e-05, # fraction susceptible to bite
                alpha   = 1.904753e-03, # rate of bites
                beta    = 2.233049e+00, # between-host infection
                gamma_b = 2.293953e-01, # E to I
                gamma_p = 2.863875e-01, # E to I
                delta_b = 2.626527e-01, # recovery + death
                delta_p = 3.413748e-01, # recovery + death
                epsilon = 3.204494e-02, # I_b to I_p
                kk      = 1.105443e+00, # Imperfect intervention
                N       = 25570895     # population size
                )

S_b_0 <- 1651
S_p_0 <- 25569244

state <- c(Sb = S_b_0, Sy = S_p_0, Eb = 0, Ey =  0, Ib = 1, Ip = 1)

times <- seq(0, 70 - 1, by = 0.1)

# Define the flea function for the flea population
flea <- function (t){
  x <- A + B * sin((pi/180.) * t) + C * cos((pi/180.) * t )

  return (x)
}
# Define the intervention functions of human intervention on the disease
intervention_p <- function (t){
  return (1 - 1/(as.numeric(parameters["kk"]) + exp(tau_p - t)))
}
intervention_b <- function (t){
  return (1 - 1/(as.numeric(parameters["kk"]) + exp(tau_b - t)))
}

```
```{r Deterministic model}
# Define the function for the model of the deterministic data runs
deterministic_model <- function(t, y, parms){
  with(as.list(c(parms, y, t)),{
    flea_t <- flea(t)
    intervention_b1 <- intervention_b(t)
    intervention_p1 <- intervention_p(t)

    delta.Sb <- -alpha * flea_t * intervention_b1 * Sb - -beta * intervention_p1 * Sb * Ip/N
    delta.Sy <- -beta * intervention_p1 * Sy * Ip/N
    delta.Eb <- alpha * flea_t * intervention_b1 * Sb - gamma_b * Eb
    delta.Ey <- beta * intervention_p1 * (Sb+Sy) * Ip/N - gamma_p * Ey
    delta.Ib <- gamma_b * Eb - epsilon * Ib - delta_b * Ib
    delta.Ip <- gamma_p * Ey + epsilon * Ib - delta_p * Ip
    print(delta.Sb)


        return(list(c(delta.Sb, delta.Sy, delta.Eb, delta.Ey, delta.Ib, delta.Ip)))
       })
}

# Run the ode model and save the output to the out variable
out <- ode(times = times, y = state,   parms = parameters, func = deterministic_model, method = "lsoda", atol = 1e-10, rtol = 1e-10)

plot(as.data.frame(out)$Eb, type = "l", xlab = "Time in Days", ylab = "Bubonice indices", col = "blue", ylim = c(0, 30))
plot(as.data.frame(out)$Ey, type = "l", xlab = "Time in Days", ylab = "Puneonic indices", col = "blue", ylim = c(0, 120))
plot(out)
plot(data$Ip)

```
In the graph(s), we can see the plot of the deterministic data model.
```{r stochastic}
state_sto <- c(state[['Sb']],state[['Sy']],state[['Eb']],state[['Ey']],state[['Ib']],state[['Ip']])
D <- 6
# Vectorize the flea & intervention functions
# v_flea <- vectorize(flea)
# v_itv_p <- vectorize(intervention_p)
# v_itv_b <- vectorize(intervention_b)

# Save all the flea/intervention data based on each piece of time
f_irf <- flea(times)
f_itv_b <- intervention_b(times)
f_itv_p <- intervention_p(times)

unit_vector <- function (i, D){
  out <- rep(0, D)
  out[i] <- 1
  return (out)
}



transition_rates <- function (state, f_irf, f_itv_p, f_itv_b) {

  return (array(c(parameters[["alpha"]] * f_irf * f_itv_b * state[["Sb"]],
                parameters[["beta"]] * f_itv_p * state[["Sb"]] * state[["Ip"]]/parameters[["N"]],
                parameters[["beta"]] * f_itv_p * state[["Sy"]] * state[["Ip"]]/parameters[["N"]],
                parameters[["gamma_b"]] * state[["Eb"]],
                parameters[["gamma_p"]] * state[["Ey"]],
                parameters[["delta_b"]] * state[["Ib"]],
                parameters[["delta_p"]] * state[["Ip"]],
                parameters[["epsilon"]] * state[["Ib"]])))

}

nu <- array(c(unit_vector(2, D) - unit_vector(0, D), unit_vector(3, D) - unit_vector(0, D),
               unit_vector(3, D) - unit_vector(1, D), unit_vector(4, D) - unit_vector(2, D),
               unit_vector(5, D) - unit_vector(3, D), -unit_vector(4, D), -unit_vector(5, D),
               unit_vector(5, D) - unit_vector(4, D)))

change_func <- function(n, firings, nu) {
  prechange <- rep(0,6)
  for (i in seq(1,n)) {
    prechange <- prechange + (firings[i]*nu[i])

  }
  return( prechange)
}

plague_tau <- function (parms, t, f_irf, f_itv_p, f_itv_b, nu) {
  x_t <- data.frame(matrix(ncol = 6, nrow = 0))
  colnames(x_t)<- c("Sb", 'Sy', 'Eb', 'Ey', 'Ib', "Ip")
  for (i in seq(0,length(t))){
    x_t[nrow(x_t) + 1,] <- state_sto
    rates <- transition_rates(state, f_irf[i], f_itv_p[i], f_itv_b[i])
    n <- length(rates)
    firings <- rpois(rates, rep(1, n))*0.1
    change <- change_func(n, firings, nu)
    print(change)
    state_sto <<- state_sto + change

  }
  x_t <- (x_t)
  return(x_t)
}

# Set the seed for the random number generator
set.seed(10042018)
runs <- 20

I_b <- da
I_P <- c()

runs <- 1
repeat {
  y <- plague_tau(parms, times, f_irf, f_itv_p, f_itv_b, nu)

  I_b <- y$Ib
  I_P <- y$Ip

  if(runs > 20) {
      break
   }
  runs <- runs +1
}

```

# Conclusion


# Discussion

While translating the python model to R.
We found that there are some minor internal rounding differences when calculating floats in R and python,
these differences where isolated and removed, simply by comparing each outcome of the r and python equation and
the outcome of each equation used for calculating the delta value was exactly the same.
Furthermore, we rewrote the R code back to python code to see if there were any differences or mistakes in our code, compared to the document's code.
In the end we only managed to find one mistake between the code. Although even after fixing the error, the results in R are different from python, even between our own code.

This let to that we found that DeSolve ode function and the scipy odeint have minor differences in calculating the deltas, that with a given time create a snowball effect.
This results in the R model having a higher peak curve compared to the Python 3 model.

Also, while converting the stochastic model from python to R.
It was apparent that we had to fully rewrite the code that they have used as it wasn't easily use-able in R itself.
This has lead to some complications in writing the code in an R-styled manner.

# Sources

https://www.sciencedirect.com/science/article/pii/S1755436518300070

